Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-17 talk/dls2012/paper.tex
use only eight characters for the hash
some improvements proposed by David

We have implemented the scheme in PyPy's tracing JIT compiler,
where it can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance into the ballpark of static language compilers.
\end{abstract}

@@

One disadvantage of this simplicity is that such forward-passing
optimizers ignore the only bit of control flow they have available, which is
the fact that most traces actually represent loops. Making use of this
information is necessary to perform optimizations that take the whole loop into
account, such as loop-invariant code
motion or optimizations that improve across several iterations of the loop.
Having to deal with this property of traces complicates the optimization passes,
as a more global view of a trace needs to be considered when optimizing.

@@

\begin{itemize}
    \item \lstinline{new} creates a new object.
    \item \lstinline{get} reads an attribute of an object.
    \item \lstinline{set} writes to an attribute of an object.
    \item \lstinline{guard_class} is a precise type check. It typically precedes
    an (inlined) method call and is followed by the trace of the called method.
    The type that the guard checks for is the one that the variable had during
    tracing.
\end{itemize}

Method calls in the trace are preceded by a \lstinline{guard_class}
operation, to check that the class of the receiver is the same as the one that
was observed during tracing.\footnote{\lstinline{guard_class}
performs a precise
class check, not checking for subclasses.} These guards make the trace specific
to the situation where \lstinline{y} is really a \lstinline{BoxedInteger}. When
the trace is turned into machine code and afterwards executed with
\lstinline{BoxedFloat}, the
first \lstinline{guard_class} instruction will fail and execution will continue
using the interpreter.

@@

Before a trace is passed to the backend compiling it into machine code
it is optimized to achieve better performance.
One goal of that is to move 
operations out of the loop making them executed only once
and not every iteration. This can be achieved by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimizations of interest only a few
additional details have to be considered when they are combined with loop peeling. These are
described below by explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

Loop peeling is achieved by appending an copy of the traced iteration at
the end of itself. See Figure~\ref{fig:overview} for an illustration.
The first part (called \emph{preamble}) finishes with a jump to the second part
(called the \emph{peeled loop}). The second part finishes with a jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintain the SSA-property.

When peeling the loop, no assumptions are made that the preamble is
the \emph{first} iteration when later executing the loop. The preamble stays
general enough to correspond to any iteration of the loop.
However, the peeled loop can then be optimized using the assumption that a
previous iteration (the preamble) has been executed already.

@@

When applying optimizations to this two-iteration trace
some care has to taken as to how the arguments of the two
\lstinline{jump} operations and the input arguments of the peeled loop are
treated. It has to be ensured that the peeled loop stays a proper
trace in the sense that the operations within it only operates on
variables that are either among its input arguments 
or produced within the peeled loop. To ensure this we need
to introduce a bit of formalism.

@@

No special concern needs to be taken when implementing redundant
guard removal together with loop peeling. The guards from
the preamble might make the guards of the peeled loop
redundant and thus removed. Therefore one effect of combining redundant
guard removal with loop peeling is that loop-invariant guards are moved out of the
loop. The peeled loop of the example reduces to the trace in Figure~\ref{fig:guard-trace}.

@@

If a pure operation appears more than once in the trace with the same input
arguments, it only needs to be executed the first time and then the result
can be reused for all other appearances. RPython's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value.\footnote{We perform a type-based alias analysis to know which
writes can affect which reads~\cite{diwan_type-based_1998}. In addition writes
on newly allocated objects can never change the value of old existing ones.}

@@

It is interesting to note that the described approach automatically deals with
implicit control dependencies correctly, whereas in other approaches this needs
to be carefully programmed in. A commonly used example for a control dependency
is a division operation that needs to be preceded by a check for the second
argument being 0. In a trace, such a check would be done with a guard. The
division operation must not be moved before that guard, and indeed, this is
never done. If the division is loop invariant, the result computed in copy of
the division operation in the preamble is reused. This division operation is
preceded by a copy of the non-null guard, which ensures that it can be executed
correctly.

@@

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
attributes.\footnote{This is sometimes called \emph{scalar replacement}~\cite{kotzmann_escape_2005}.}
If some of the attributes are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed from $\hat J$ using the map $m$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $J$ is replaced by $\hat J$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to
the trace in Figure~\ref{fig:virtual-trace}.

@@

If all the optimizations presented above are applied, the resulting loop looks
as in Figure~\ref{fig:opt-trace}.
The resulting optimized peeled loop consists of a single integer addition
only. That is it will become type-specialized to the types of the
variables \lstinline{step} and \lstinline{y}, and the overhead of
using boxed values is removed.

@@

\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|r|}
\hline
 & CPython & PyPy  & PyPy & LuaJIT & LuaJIT & GCC \\
 &         & no LP &      & no LP  &        & -O3 \\
\hline
FFT(1024,32768) & 469.07 & 20.83 $\pm$ 0.039 & 12.73 $\pm$ 0.029 & 4.45 $\pm$ 0.019 & 2.74 $\pm$ 0.021 & 1.40 $\pm$ 0.082\\
\hline
FFT(1048576,2) & 58.93 & 4.12 $\pm$ 0.020 & 2.05 $\pm$ 0.007 & 1.25 $\pm$ 0.019 & 1.07 $\pm$ 0.050 & 0.83 $\pm$ 0.044\\
\hline
LU(100,4096) & 1974.14 & 32.22 $\pm$ 0.281 & 13.39 $\pm$ 0.063 & 8.57 $\pm$ 0.018 & 1.52 $\pm$ 0.010 & 1.33 $\pm$ 0.070\\
\hline
LU(1000,2) & 955.31 & 14.98 $\pm$ 0.436 & 5.99 $\pm$ 0.416 & 4.00 $\pm$ 0.018 & 0.67 $\pm$ 0.014 & 0.65 $\pm$ 0.077\\
\hline
MonteCarlo(268435456) & 618.89 & 20.60 $\pm$ 0.097 & 15.33 $\pm$ 0.163 & 3.92 $\pm$ 0.013 & 2.82 $\pm$ 0.010 & 1.69 $\pm$ 0.096\\
\hline
SOR(100,32768) & 1458.12 & 8.24 $\pm$ 0.002 & 2.66 $\pm$ 0.002 & 2.02 $\pm$ 0.011 & 1.31 $\pm$ 0.010 & 1.76 $\pm$ 0.088\\
\hline
SOR(1000,256) & 1210.45 & 6.48 $\pm$ 0.007 & 2.10 $\pm$ 0.005 & 1.63 $\pm$ 0.006 & 1.08 $\pm$ 0.014 & 1.49 $\pm$ 0.042\\
\hline
SparseMatMult(1000,5000,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.69 $\pm$ 0.033 & 4.49 $\pm$ 0.036 & 1.84 $\pm$ 0.061\\
\hline
SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.019 & 2.43 $\pm$ 0.031 & 1.20 $\pm$ 0.053\\
\hline
\hline
conv3(1e6) & 49.20 & 1.13 $\pm$ 0.043 & 0.51 $\pm$ 0.008 & 0.70 $\pm$ 0.009 & 0.18 $\pm$ 0.009 & 0.60 $\pm$ 0.064\\
\hline
conv3x3(1000,1000) & 138.95 & 0.70 $\pm$ 0.007 & 0.20 $\pm$ 0.009 & 0.22 $\pm$ 0.009 & 0.15 $\pm$ 0.010 & 0.17 $\pm$ 0.079\\
\hline
dilate3x3(1000,1000) & 137.52 & 4.35 $\pm$ 0.014 & 3.91 $\pm$ 0.037 & 0.22 $\pm$ 0.008 & 0.16 $\pm$ 0.010 & 0.17 $\pm$ 0.061\\
\hline
sobel(1000,1000) & 104.02 & 0.49 $\pm$ 0.009 & 0.21 $\pm$ 0.004 & 0.37 $\pm$ 0.014 & 0.24 $\pm$ 0.017 & 0.17 $\pm$ 0.061\\
\hline
sqrt(float) & 14.99 & 1.37 $\pm$ 0.001 & 0.89 $\pm$ 0.000 & 1.06 $\pm$ 0.010 & 0.83 $\pm$ 0.014 & 0.85 $\pm$ 0.088\\
\hline
sqrt(int) & 13.91 & 3.22 $\pm$ 0.033 & 2.65 $\pm$ 0.001 & - & - & 1.25 $\pm$ 0.053\\
\hline
sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & 4.00 $\pm$ 0.040 & 1.47 $\pm$ 0.014 & 1.34 $\pm$ 0.061\\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark results in seconds with 95\% confidence intervals. The leftmost column gives the
name of each benchmark and the values of the benchmark parameters used. The different benchmarks and the meaning of their parameters are described in Section~\ref{sec:benchmarks}.}
\end{figure*}

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{benchmarks/result.pdf}
\label{fig:benchmarks_plot}
\caption{Benchmark results normalized with the runtime of the C version. The CPython results have been omitted to make the plot readable.}
\end{center}
\end{figure}

The Python interpreter of the RPython framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python, C and Lua and their
runtimes are compared in Figuare~\ref{fig:benchmarks_plot} and Figure~\ref{fig:benchmarks}.\footnote{
    The benchmarks and the scripts to run them can be found in the repository for this paper:
    \texttt{https://bitbucket.org/pypy/extradoc/src/ tip/talk/dls2012/benchmarks}
}
For benchmarks using larger Python applications the times are unaffected or
slightly improved by the loop optimization of this paper.

@@

\item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop
is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_{n-2}\right)$ from a vector
${\bf a} = \left(a_1, \cdots, a_n\right)$ and a kernel ${\bf k} = \left(k_1, k_2, k_3\right)$ using 
$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n-2$. Both the output vector, $\bf b$, 
and the input vectors, $\bf a$ and $\bf k$, are allocated prior to running the benchmark. It is executed 
with $n=10^5$.
%\item {\bf conv5}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
%${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
hardcoded into the implementation making the benchmark consist of a single loop too.
\item {\bf conv3x3}$\left(n,m\right)$: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays. It is implemented as two nested loops that iterates over the elements of the 
$m\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
${\bf A} = \left(a_{i,j}\right)$ and a kernel ${\bf K} = \left(k_{i,j}\right)$ using $b_{i,j} = $
\begin{equation}
  \label{eq:convsum}
  \begin{array}{lclclc}
    k_{3,3} a_{i-1,j-1} &+& k_{3,2} a_{i-1,j} &+& k_{3,1} a_{i-1,j+1} & + \\
    k_{2,3} a_{i,j-1}   &+& k_{2,2} a_{i,j}   &+& k_{2,1} a_{i,j+1}   & + \\
    k_{1,3} a_{i+1,j-1} &+& k_{1,2} a_{i+1,j} &+& k_{1,1} a_{i+1,j+1}  \\
  \end{array}
\end{equation}
for $2 \leq i \leq m-1$ and $2 \leq j \leq n-1$.
The memory for storing the matrices are again allocated outside the benchmark and $(n,m)=(1000,1000)$ 
 was used.
\item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the terms in Equation~\ref{eq:convsum}, the maximum over those terms is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}$\left(n\right)$: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. A Sobel x-derivative, $D_x$, of a $n \times n$ image, ${I}$, is formed
by convolving ${I}$ with
\begin{equation}
  {K} = \left(
  \begin{array}{ccc}
    -1 & 0 & 1 \\
    -2 & 0 & 2 \\
    -1 & 0 & 1 \\
  \end{array}
  \right) ,
\end{equation}
and a Sobel y-derivative, $D_y$, is formed convolving $I$ with $K^\top$. The gradient magnitude is 
then formed for each pixel independently by $\sqrt{D_x^2 + D_y^2}$. The two convolutions and the pixelwise
magnitude calculation are combined in the implementation of this benchmark and calculated in a single pass over
the input image. This single pass consists of two nested loops with a somewhat larger amount of calculations 
performed each iteration as compared to the other benchmarks.
\end{itemize}

@@

Benchmarks were run on Intel Xeon X5680 @3.33GHz with 12M cache and 16G of RAM
using Ubuntu Linux 11.4 in 64bit mode.
The machine was otherwise unoccupied. We use the following software
for benchmarks:

\begin{itemize}
\item PyPy 1.9
\item CPython 2.7.1
\item GCC 4.5.2 shipped with Ubuntu 11.4
\item LuaJIT 2.0 beta, git head of August 15, 2012, commit ID 0dd175d9e711f039c663d35e96c149b705bcf450
\end{itemize}

We run GCC with -O3 -march=native, disabling the
automatic loop vectorization. In all cases, SSE2 instructions were used for
floating point operations.
We also run PyPy and LuaJIT with loop peeling optimization and without (but otherwise
identical).

For PyPy and LuaJIT 10 iterations were run, prefaced with 3 iterations for warming up.
Due to benchmarks taking large amounts of time on CPython, only one run
was performed.
For GCC 5 iterations
were run. In all cases, the standard deviation is very low, making benchmarks
very well reproducible.

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than CPython. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}, the main improvement for
these concrete benchmarks come from the allocation removal/unboxing
optimization.

@@

Loop invariant code motion has been part of early compilers in the 1960s and
1970s~\cite{allen_catalogue_1971}. A common approach for achieving loop invariant
code motion is to perform partial redundancy elimination. The
approach was first proposed by Morel and Renvoise~\cite{morel_global_1979}. It
involves solving data flow problems of bidirectional data flow
equations. After improvements~\cite{chow_portable_1984,
dhamdhere_practical_1991} this approach was followed by the work of Knoop
et.al.~\cite{knoop_lazy_1992} who cleanly separated the problem into a backward
and forward data flow analysis. Implementing partial redundancy elimination in
compilers that use SSA form \cite{chow_new_1997} simplified the algorithms
because no iterative data flow analysis is needed any more.

@@

The current approach still has some limitations which we plan to address in the
future. In particular loop peeling works poorly in combination with trace
trees~\cite{gal_incremental_2006} or trace stitching~\cite{gal_trace-based_2009}.
The side exits attached guards that fail often
currently have to jump to the preamble which makes loops with several equally
common paths less efficient than they could be.

>>>>

We have implemented the scheme in RPython's tracing JIT compiler. PyPy's Python
JIT executing simple numerical kernels can become up to two times faster,
bringing the performance into the ballpark of static language compilers.
\end{abstract}

@@

One disadvantage of this simplicity is that such forward-passing
optimizers ignore the only bit of control flow they have available, which is
the fact that most traces actually represent loops. Making use of this
information is necessary to perform optimizations that take the whole loop into
account, such as loop-invariant code
motion or optimizations that improve several iterations of the loop.
Having to deal with this property of traces complicates the optimization passes,
as a more global view of a trace needs to be considered when optimizing.

@@

\begin{itemize}
    \item \lstinline{new} creates a new object.
    \item \lstinline{get} reads an attribute of an object.
    \item \lstinline{set} writes to an attribute of an object.
    \item \lstinline{guard_class} is a precise type check, not checking for subclasses.
\end{itemize}

Inlined method calls in the trace are preceded by a \lstinline{guard_class}
operation, to check that the class of the receiver is the same as the one that
was observed during tracing. These guards make the trace specific
to the situation where \lstinline{y} is really a \lstinline{BoxedInteger}. When
the trace is turned into machine code and afterwards executed with
\lstinline{BoxedFloat}, the
first \lstinline{guard_class} instruction will fail and execution will continue
using the interpreter.

@@

Before a trace is compiled to machine code by the backend,
it is optimized to achieve better performance.
One goal of that is to move 
operations out of the loop to execute them only once
and not every iteration. This can be achieved by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimizations of interest only a few
additional details have to be considered when they are combined with loop peeling. These are
described below by explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

Loop peeling is achieved by appending a copy of the traced iteration at
the end of itself. See Figure~\ref{fig:overview} for an illustration.
The first part (called \emph{preamble}) finishes with a jump to the second part
(called the \emph{peeled loop}). The second part finishes with a jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintain the SSA-property.

When peeling the loop, no assumptions are made that the preamble is
the \emph{first} iteration, when later executing the loop. The preamble stays
general enough to correspond to any iteration of the loop.
However, the peeled loop can then be optimized using the assumption that a
previous iteration (the preamble) has been executed already.

@@

When applying optimizations to this two-iteration trace
some care has to taken as to how the arguments of the two
\lstinline{jump} operations and the input arguments of the peeled loop are
treated. It has to be ensured that the peeled loop stays a proper
trace in the sense that the operations within it only operate on
variables that are either among its input arguments 
or produced within the peeled loop. To ensure this we need
to introduce a bit of formalism.

@@

Redundant guard removal removes guards that are implied by other guards earlier
in the trace. The most common case is the removal of a guard that has already
appeared.
No special concern needs to be taken when implementing redundant
guard removal together with loop peeling. The guards from
the preamble might make the guards of the peeled loop
redundant and thus removed. Therefore one effect of combining redundant
guard removal with loop peeling is that loop-invariant guards are moved out of the
loop. The peeled loop of the example reduces to the trace in Figure~\ref{fig:guard-trace}.

@@

If a pure operation appears more than once in the trace with the same input
arguments, it only needs to be executed the first time and then the result
can be reused for all other appearances. This is achieved by common
subexpression elimination. RPython's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value.\footnote{We perform a type-based alias analysis to know which
writes can affect which reads~\cite{diwan_type-based_1998}. In addition writes
on newly allocated objects can never change the value of old existing ones.}

@@

It is interesting to note that the described approach deals correctly with
implicit control dependencies, whereas in other approaches this needs
to be carefully programmed in. A commonly used example for a control dependency
is a division operation that needs to be preceded by a check for the second
argument being 0. In a trace, such a check would be done with a guard. The
division operation must not be moved before that guard, and indeed, this is
never done. If the division is loop invariant, the result computed by the copy of
the division operation in the preamble is reused. This division operation is
preceded by a copy of the guard that checks that the second argument is not 0,
which ensures that the division can be executed correctly.
Such control dependencies are common in traces produced by dynamic languages.
Reading a field out of an object is often preceded by checking the type of the
object.

@@

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
attributes.\footnote{This is sometimes called \emph{scalar replacement}~\cite{kotzmann_escape_2005}.}
If some of the attributes are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if every
concrete variable of the jump arguments is also exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed from $\hat J$ using the map $m$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $J$ is replaced by $\hat J$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to
the trace in Figure~\ref{fig:virtual-trace}.

@@

If all the optimizations presented above are applied, the resulting loop looks
as in Figure~\ref{fig:opt-trace}.
The resulting optimized peeled loop consists of a single integer addition. That
is it will become type-specialized to the types of the
variables \lstinline{step} and \lstinline{y}, and the overhead of
using boxed values is removed.

@@

\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|r|}
\hline
 & CPython & PyPy  & PyPy & LuaJIT & LuaJIT & GCC \\
 &         & no LP &      & no LP  &        & -O3 \\
\hline
FFT(1024,32768) & 469.07 & 20.83 $\pm$ 0.039 & 12.73 $\pm$ 0.029 & 4.45 $\pm$ 0.019 & 2.74 $\pm$ 0.021 & 1.40 $\pm$ 0.082\\
\hline
FFT(1048576,2) & 58.93 & 4.12 $\pm$ 0.020 & 2.05 $\pm$ 0.007 & 1.25 $\pm$ 0.019 & 1.07 $\pm$ 0.050 & 0.83 $\pm$ 0.044\\
\hline
LU(100,4096) & 1974.14 & 32.22 $\pm$ 0.281 & 13.39 $\pm$ 0.063 & 8.57 $\pm$ 0.018 & 1.52 $\pm$ 0.010 & 1.33 $\pm$ 0.070\\
\hline
LU(1000,2) & 955.31 & 14.98 $\pm$ 0.436 & 5.99 $\pm$ 0.416 & 4.00 $\pm$ 0.018 & 0.67 $\pm$ 0.014 & 0.65 $\pm$ 0.077\\
\hline
MonteCarlo(268435456) & 618.89 & 20.60 $\pm$ 0.097 & 15.33 $\pm$ 0.163 & 3.92 $\pm$ 0.013 & 2.82 $\pm$ 0.010 & 1.69 $\pm$ 0.096\\
\hline
SOR(100,32768) & 1458.12 & 8.24 $\pm$ 0.002 & 2.66 $\pm$ 0.002 & 2.02 $\pm$ 0.011 & 1.31 $\pm$ 0.010 & 1.76 $\pm$ 0.088\\
\hline
SOR(1000,256) & 1210.45 & 6.48 $\pm$ 0.007 & 2.10 $\pm$ 0.005 & 1.63 $\pm$ 0.006 & 1.08 $\pm$ 0.014 & 1.49 $\pm$ 0.042\\
\hline
SparseMatMult(1000,5000,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.69 $\pm$ 0.033 & 4.49 $\pm$ 0.036 & 1.84 $\pm$ 0.061\\
\hline
SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.019 & 2.43 $\pm$ 0.031 & 1.20 $\pm$ 0.053\\
\hline
\hline
conv3(1e6) & 49.20 & 1.13 $\pm$ 0.043 & 0.51 $\pm$ 0.008 & 0.70 $\pm$ 0.009 & 0.18 $\pm$ 0.009 & 0.60 $\pm$ 0.064\\
\hline
conv3x3(1000,1000) & 138.95 & 0.70 $\pm$ 0.007 & 0.20 $\pm$ 0.009 & 0.22 $\pm$ 0.009 & 0.15 $\pm$ 0.010 & 0.17 $\pm$ 0.079\\
\hline
dilate3x3(1000,1000) & 137.52 & 4.35 $\pm$ 0.014 & 3.91 $\pm$ 0.037 & 0.22 $\pm$ 0.008 & 0.16 $\pm$ 0.010 & 0.17 $\pm$ 0.061\\
\hline
sobel(1000,1000) & 104.02 & 0.49 $\pm$ 0.009 & 0.21 $\pm$ 0.004 & 0.37 $\pm$ 0.014 & 0.24 $\pm$ 0.017 & 0.17 $\pm$ 0.061\\
\hline
sqrt(float) & 14.99 & 1.37 $\pm$ 0.001 & 0.89 $\pm$ 0.000 & 1.06 $\pm$ 0.010 & 0.83 $\pm$ 0.014 & 0.85 $\pm$ 0.088\\
\hline
sqrt(int) & 13.91 & 3.22 $\pm$ 0.033 & 2.65 $\pm$ 0.001 & - & - & 1.25 $\pm$ 0.053\\
\hline
sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & 4.00 $\pm$ 0.040 & 1.47 $\pm$ 0.014 & 1.34 $\pm$ 0.061\\
\hline
\end{tabular}
}
\end{center}
\caption{Benchmark results in seconds with 95\% confidence intervals. The leftmost column gives the
name of each benchmark and the values of the benchmark parameters used. The different benchmarks and the meaning of their parameters are described in Section~\ref{sec:benchmarks}.}
\label{fig:benchmarks}
\end{figure*}

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{benchmarks/result.pdf}
\end{center}
\caption{Benchmark results normalized to the runtime of the C version. The CPython results have been omitted to make the plot readable.}
\label{fig:benchmarks_plot}
\end{figure}

The Python interpreter of the RPython framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python, C and Lua and their
runtimes are compared in Figure~\ref{fig:benchmarks_plot} and Figure~\ref{fig:benchmarks}.\footnote{
    The benchmarks and the scripts to run them can be found in the repository for this paper:
    \texttt{https://bitbucket.org/pypy/extradoc/src/ tip/talk/dls2012/benchmarks}
}
For benchmarks using larger Python applications the times are unaffected or
only slightly improved by the loop optimization of this paper.

@@

\item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop
is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_{n-2}\right)$ from a vector
${\bf a} = \left(a_1, \cdots, a_n\right)$ and a kernel ${\bf k} = \left(k_1, k_2, k_3\right)$ using 
$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n-2$. Both the output vector, $\bf b$, 
and the input vectors, $\bf a$ and $\bf k$, are allocated prior to running the benchmark. It is executed 
with $n=10^5$.
%\item {\bf conv5}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
%${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
hardcoded into the implementation making the benchmark consist of a single loop too.
\item {\bf conv3x3}$\left(n,m\right)$: two-dimensional convolution with a kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays. It is implemented as two nested loops that iterates over the elements of the 
$m\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
${\bf A} = \left(a_{i,j}\right)$ and a kernel ${\bf K} = \left(k_{i,j}\right)$ using $b_{i,j} = $
\begin{equation}
  \label{eq:convsum}
  \begin{array}{lclclc}
    k_{3,3} a_{i-1,j-1} &+& k_{3,2} a_{i-1,j} &+& k_{3,1} a_{i-1,j+1} & + \\
    k_{2,3} a_{i,j-1}   &+& k_{2,2} a_{i,j}   &+& k_{2,1} a_{i,j+1}   & + \\
    k_{1,3} a_{i+1,j-1} &+& k_{1,2} a_{i+1,j} &+& k_{1,1} a_{i+1,j+1}  \\
  \end{array}
\end{equation}
for $2 \leq i \leq m-1$ and $2 \leq j \leq n-1$.
The memory for storing the matrices are again allocated outside the benchmark and $(n,m)=(1000,1000)$ 
 was used.
\item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with a kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the terms in Equation~\ref{eq:convsum}, the maximum over those terms is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}$\left(n\right)$: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. A Sobel x-derivative, $D_x$, of a $n \times n$ image, ${I}$, is formed
by convolving ${I}$ with
\begin{equation}
  {K} = \left(
  \begin{array}{ccc}
    -1 & 0 & 1 \\
    -2 & 0 & 2 \\
    -1 & 0 & 1 \\
  \end{array}
  \right) ,
\end{equation}
and a Sobel y-derivative, $D_y$, is formed convolving $I$ with $K^\top$. The gradient magnitude is 
then formed for each pixel independently by $\sqrt{D_x^2 + D_y^2}$. The two convolutions and the pixelwise
magnitude calculation are combined in the implementation of this benchmark and calculated in a single pass over
the input image. This single pass consists of two nested loops with a somewhat larger amount of calculations 
performed each iteration as compared to the other benchmarks.
\end{itemize}

@@

Benchmarks were run on Intel Xeon X5680 @3.33GHz with 12M cache and 16G of RAM
using Ubuntu Linux 11.4 in 64bit mode.
The machine was otherwise unoccupied. We used the following software
for benchmarks:

\begin{itemize}
\item PyPy 1.9
\item CPython 2.7.1
\item GCC 4.5.2 shipped with Ubuntu 11.4
\item LuaJIT 2.0 beta, git head of August 15, 2012, commit ID 0dd175d9
\end{itemize}

We ran GCC with -O3 -march=native, disabling the
automatic loop vectorization. In all cases, SSE2 instructions were used for
floating point operations.
We also ran PyPy and LuaJIT with loop peeling optimization and without (but otherwise
identical).

For PyPy and LuaJIT, 10 iterations were run, prefaced with 3 iterations for warming up.
Due to benchmarks taking large amounts of time on CPython, only one run
was performed.
For GCC, 5 iterations
were run. In all cases, the standard deviation is very low, making benchmarks
very well reproducible.

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than CPython. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}, the main improvement for
these concrete benchmarks comes from the allocation removal/unboxing
optimization.

@@

Loop invariant code motion has been part of early compilers since the
1960s~\cite{allen_catalogue_1971}. A common approach for achieving loop
invariant
code motion is to perform partial redundancy elimination. The
approach was first proposed by Morel and Renvoise~\cite{morel_global_1979}. It
involves solving data flow problems of bidirectional data flow
equations. After improvements~\cite{chow_portable_1984,
dhamdhere_practical_1991} this approach was followed by the work of Knoop
et.al.~\cite{knoop_lazy_1992} who cleanly separated the problem into a backward
and forward data flow analysis. Implementing partial redundancy elimination in
compilers that use SSA form~\cite{chow_new_1997} simplified the algorithms,
because no iterative data flow analysis was needed any more.

@@

The described approach still has some limitations which we plan to address in the
future. In particular loop peeling works poorly in combination with trace
trees~\cite{gal_incremental_2006} or trace stitching~\cite{gal_trace-based_2009}.
The side exits attached to guards that fail often
currently have to jump to the preamble.
